\documentclass[runningheads]{llncs}

% ── Core packages ──────────────────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{tabularx}
\usepackage[table,dvipsnames]{xcolor}
\usepackage{pgfplots}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows.meta,positioning,fit,backgrounds}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[hidelinks]{hyperref}
\usepackage{url}
\usepackage{natbib}
\usepackage{microtype}      % better justification
\usepackage{float}
\usepackage{placeins}       % \FloatBarrier
\pgfplotsset{compat=1.18}
\def\UrlFont{\rmfamily}

% ── Spacing helpers ─────────────────────────────────────────────────────────────
\setlength{\textfloatsep}{8pt plus 2pt minus 2pt}
\setlength{\floatsep}{6pt plus 2pt minus 2pt}
\setlength{\intextsep}{6pt plus 2pt minus 2pt}
\setlength{\abovecaptionskip}{4pt}
\setlength{\belowcaptionskip}{2pt}

% ── Colour palette for heatmap ──────────────────────────────────────────────────
\definecolor{hBlue}{RGB}{198,219,239}
\definecolor{hGreen}{RGB}{186,228,179}
\definecolor{hYellow}{RGB}{254,217,142}
\definecolor{hOrange}{RGB}{253,174,107}
\definecolor{hRed}{RGB}{240,159,140}
\definecolor{hGray}{RGB}{220,220,220}

% ───────────────────────────────────────────────────────────────────────────────
\begin{document}

\title{Constraints and Heuristics: Leveraging Dataset Metadata for Efficient AutoML}
\titlerunning{Constraints and Heuristics for Efficient AutoML}

\author{Harsh Kakadiya\inst{1} \and
        Juli Kyada\inst{1}    \and
        Kaushal Savaliya\inst{1} \and
        Jalpesh Vasa\inst{1}}
\authorrunning{Harsh Kakadiya et al.}

\institute{Department of Artificial Intelligence and Machine Learning,\\
           Chandubhai S.\ Patel Institute of Technology (CSPIT),\\
           Faculty of Technology and Engineering,\\
           Charotar University of Science and Technology (CHARUSAT),\\
           Changa, Gujarat, India\\
           \email{harshkakadiya128@gmail.com, julikyada293@gmail.com, kaushalsavaliya2627@gmail.com, jalpeshvasa.it@charusat.ac.in}}

\maketitle

% ══════════════════════════════════════════════════════════════════════════════
\begin{abstract}
This work introduces MetaFlow, a lightweight AutoML pipeline selection framework based on metadata, which automatically prunes and re-ranks candidate pipelines before training such that the available information about the data sample's size, types of features and their potential imbalance, their cardinality and their outlier density is used deterministically without having to train each different pipeline separately. MetaFlow uses a small number of rules with high semantic content that are easily interpretable to avoid evaluating clearly inappropriate learners and can achieve compute savings by pruning the search space reported here as $\sim$67\%. Its main contribution is a practical ``grey-box'' strategy that executes a targeted search for efficiency instead of searching through the breadth of possible options using human-interpretable constraints and prioritization.

The important note is that the pruning and warm-start tips from the perspective of Data analysis are proposed giving it as an alternate inexpensive approach to modern AutoML toolkits like Auto-sklearn, FLAML, etc. Constraints are also noted: The higher (or categorical) thresholds (such as percentage of missingness) are heuristic and data set dependent, and only representative tabular cases are studied as data sets for empirical support rather than a comprehensive benchmark. A reproducibility checklist and a benchmarking protocol are explained in order to be able to easily compare with other state of the art AutoML systems.

\keywords{AutoML \and metadata \and pipeline selection \and meta-learning \and algorithm selection \and data profiling \and preprocessing}
\end{abstract}

% ══════════════════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}

Creating robust machine learning workflows that involve data cleaning, feature engineering, algorithm selection and hyperparameter optimization, requires significant domain expertise and computational resources. The challenge is met by the AutoML systems by automating the search over pipeline configurations. The issue of hyperparameter optimisation and algorithm selection was first formalised by Auto-WEKA~\cite{thornton2013auto}, and Auto-sklearn introduced meta-learning and warm-starting~\cite{feurer2015efficient}. Modern techniques include Bayesian optimization~\cite{bergstra2011algorithms}, genetic programming~\cite{olson2016tpot} and random search~\cite{li2017hyperband} which are powerful but time-consuming, often searching many sub-optimal candidates.

A different method is based on dataset \emph{metadata} that encodes structural and statistical properties to choose, in a complementary paradigm, the search space to be warmed-up or constrained, using a framework from~\cite{leite2012survey}. Preventing the usage of inappropriate algorithms \emph{before} the training process should be possible through prior knowledge of dataset characteristics, e.g., dataset size, feature types, missingness, outliers, and class imbalance, as an automated system can be programmed to do so~\cite{wang2021flaml}. We introduce \textbf{MetaFlow}, which extracts metadata to automatically detect tasks, apply target transformations, rank models, and evaluate pipelines without exhaustive search. To guide the investigation, two main research questions are formulated: \textbf{RQ1 (Efficacy):} Can lightweight dataset metadata accurately detect the type of task and rank the learning algorithms that are appropriate? \textbf{RQ2 (Efficiency):} Does metadata-based pruning significantly impact the amount of computation compared to testing all methods?

% ══════════════════════════════════════════════════════════════════════════════
\section{Related Work}
\label{sec:related}

The CASH problem is a joint search over algorithms and hyperparameters. It was first explored and utilized by Auto-WEKA~\cite{thornton2013auto} which used a number of high-cost inferences on the black-box in this hierarchical space. TPOT~\cite{olson2016tpot} uses genetic programming to create flexible pipeline structures, but requires high computation cost in the process. Although frugal optimizers like FLAML~\cite{wang2021flaml} emphasize low-cost learners, they ignore \emph{a priori} dataset reasoning.

Meta-learning proposes algorithm selection as a supervised learning problem to previous experimental outcomes. This is boosted by Auto-sklearn~\cite{feurer2015efficient} which relies on warm starting from a meta-knowledge repository (OpenML~\cite{vanschoren2013openml}) and statistical, information theoretic and landmarking meta-features for transfer learning~\cite{leite2012survey}. This works, but needs very large, and bulky to maintain, offline repositories.

In addition to algorithm selection, metadata plays a role in determining pre-processing decisions. Target encoding is often more suitable for categoricals with large numbers of categories than one-hot encoding, and imputing strategies have more impact on downstream performance than does the choice of classification model~\cite{li2021cleanml}.

\subsection{Domain-Specific Implementations}
To conclude, domain knowledge-based machine learning techniques have shown great success in specialized problem domains such as solar power forecasting~\cite{dhaked2023power, dhaked2025exploring}, wind turbine pitch control~\cite{narayanan2024improved}, medical system cyber physical security~\cite{dhaked2025recent}, epidemic forecasting~\cite{dhaked2025prediction}, and in ranking power grid contingencies~\cite{dhaked2025ensemble}. However, such applications are based on manually designed and custom pipelines. This highlights the demand for general metadata-driven AutoML frameworks capable of automatically optimizing towards an appropriate model given the dataset characteristics.

\subsection{Research Gap and Contribution}
Optimum (general but slow) and optimum meta-learning (massive training logs) lie in a range from one another. Currently, these systems will only use metadata at initialisation time to perform black-box search and few frameworks use \emph{hard constraints} to deterministically reduce the search space. MetaFlow addresses this mismatch in a ``grey-box'' fashion: The matching of interpretable metadata is performed right to pipeline decisions, making use of a rule-based expert system to narrow down the range of incompatible candidates \emph{before} training even starts~\cite{hutter2019automated}.

Many of the recent systems, such as Auto-sklearn~\cite{feurer2015efficient} and FLAML~\cite{wang2021flaml} have associated meta-features that they use for warm starting optimization and add speed trade-offs for low cost learners, respectively. But either technique still performs many evaluations. In contrast, MetaFlow offers interpretable real-time constraints at the beginning of the pipeline that directly eliminate undesirable models following a clear and simple set of rules. MetaFlow certainly finds its niche here: It reduces the dimension of the search space with heuristics and gives the small space to the tuners that are running optimizers downstream to fine-tune the model.

% ══════════════════════════════════════════════════════════════════════════════
\section{Methodology}
\label{sec:methodology}

MetaFlow is an end-to-end system that extracts structural and statistical
metadata from a tabular dataset and uses it to drive every downstream
decision   task detection, preprocessing, model selection, and evaluation  
without exhaustive search. Figure~\ref{fig:metaflow} summarises the processing
stages.

% ── Formal problem statement ─────────────────────────────────────────────────
\subsection{Formal Problem Statement}

Let a dataset be denoted by $\mathcal{D}=(X,y)$ with $X\in\mathbb{R}^{n\times p}$
and target $y$.  The \emph{metadata extractor} is a deterministic mapping
\begin{equation}
  M:\;\mathcal{D} \mapsto m,\qquad m\in\mathcal{M},
\end{equation}
where $m$ is a structured metadata object (cardinality, missingness, $p/n$,
skewness, etc.).  A rule-based selector is a function
\begin{equation}
  R:\;\mathcal{M} \mapsto \mathcal{G},
\end{equation}
that maps metadata to a candidate model pool $\mathcal{G}=\{g_1,\dots,g_L\}$ and
associated hyperparameter grids.  A pruning operator $P$ deterministically
removes unsuitable candidates given thresholds and hard constraints:
\begin{equation}
  \mathcal{G}' = P(\mathcal{G}, m),\qquad \mathcal{G}'\subseteq\mathcal{G}.
\end{equation}
Each candidate pipeline $\pi$ in $\mathcal{G}'$ is evaluated using a
cross-validation evaluator $E$ producing a score $s$ and cost $c$:
\begin{equation}
  (s,c) = E(\pi,\mathcal{D}).
\end{equation}
The final selection chooses the pipeline that maximises a user-specified utility
($\,$e.g. test metric subject to a cost budget):
\begin{equation}
  \pi^* = \arg\max_{\pi\in\mathcal{G}'}\; U\big(E(\pi,\mathcal{D})\big).
\end{equation}

This concise formalisation highlights where metadata-driven rules intervene
prior to any model training and explicitly separates pruning, evaluation, and
selection.

% ── Pipeline figure ───────────────────────────────────────────────────────────
\begin{figure}[t]
\centering
\scalebox{0.78}{
\begin{tikzpicture}[
  node distance=0.35cm and 1.1cm,
  stage/.style={
    rectangle, rounded corners=3pt,
    minimum width=3.0cm, minimum height=0.52cm,
    text centered, draw=black!70, line width=0.45pt,
    fill=gray!6, font=\footnotesize\strut
  },
  arrow/.style={-Stealth, line width=0.6pt, draw=black!65}
]
% Left column
\node[stage] (load)     at (0,0)            {\textbf{Load Data}};
\node[stage] (generate) [below=0.58cm of load]  {\textbf{Generate Candidates}};
\node[stage] (train)    [below=0.58cm of generate] {\textbf{Train Models}};
\node[stage] (issues)   [below=0.58cm of train]   {\textbf{Detect Issues \& Plan}};
\node[stage] (explain)  [below=0.58cm of issues]  {\textbf{Generate Explanation}};
% Right column
\node[stage] (extract)  [right=of load]          {\textbf{Extract Metadata}};
\node[stage] (detect)   [below=0.58cm of extract] {\textbf{Detect Task Type}};
\node[stage] (evaluate) [below=0.58cm of detect]  {\textbf{Evaluate Performance}};
\node[stage] (select)   [below=0.58cm of evaluate]{\textbf{Select Best Pipeline}};
\node[stage] (return)   [below=0.58cm of select]  {\textbf{Return Results}};
% Arrows
\draw[arrow] (load)     -- (extract);
\draw[arrow] (extract)  -- (detect);
\draw[arrow] (detect)   -- (generate);
\draw[arrow] (generate) -- (train);
\draw[arrow] (train)    -- (evaluate);
\draw[arrow] (evaluate) -- (select);
\draw[arrow] (select)   -- (issues);
\draw[arrow] (issues)   -- (explain);
\draw[arrow] (explain)  -- (return);
\end{tikzpicture}
}
\caption{MetaFlow processing stages (left column: pipeline flow; right column: metadata-driven decisions).}
\label{fig:metaflow}
\end{figure}

Rule-based constraints encode simple, domain-informed rejection and
prioritisation rules so that clearly unsuitable algorithms are removed before
any training, saving compute and yielding human-interpretable decisions that
are easier to audit than opaque warm-start priors.

% ── Metadata Extraction ───────────────────────────────────────────────────────
\subsection{Metadata Extraction}

 The \texttt{MetadataExtractor} calculates nine groups of metadata: (1) the dimensions of the feature matrix $X$ and $y$; (2) feature types: numeric columns with $\leq \tau_{\text{cat}}=10$ unique values are classified as categorical; (3) target statistics; (4) quality profile: missing ratios per feature, duplicate-row count; (5) descriptive statistics: skewness and kurtosis per feature; (6) outlier profile: identifies values exceeding IQR boundaries; (7) cardinality profile: categorical value distributions; (8) Pearson correlations with $y$; (9) class-imbalance profile with \texttt{is\_highly\_imbalanced} as a flag when any class is with \textless{}5\% of the samples. All subsequent modules will use this metadata dictionary as their \emph{only input} so all decisions can be referred back to a property of this dataset that can be measured. Thresholds like $\theta_{\text{miss}} = 0.5$ and $\tau_{\text{cat}} = 10$ are heuristics chosen for stable, interpretable behavior on tabular data based on common ML practices. While such thresholds depend on the dataset, Section~\ref{sec:limitations} provides a discussion on how thresholds can be modified as well as a sensitivity study as a reproducibility checklist.

% ── Task Detection & Preprocessing ───────────────────────────────────────────
\subsection{Task Detection and Preprocessing}

The \texttt{TaskDetector} uses a rule chain (threshold $\tau_\text{cls}=20$ and minimum samples per class $\tau_\text{min}=10$) to determine classification vs.\ regression in $y$; assuming: non-numeric $y \Rightarrow$ classification (1.0); $|y_u|=2 \Rightarrow$ binary classification (0.95); integer $y$ with $|y_u|<\tau_\text{cls} \Rightarrow$ multiclass (0.90); $|y_u|>0.5n \Rightarrow$ regression (0.95).

Before training, the \texttt{DataPreprocessor} drops features with missing ratios ${>}\theta_{\text{miss}}=0.5$, imputes missing values (median for numeric, mode for categorical), and removes rows with missing targets. Each scikit-learn \texttt{Pipeline} includes a \texttt{ColumnTransformer} to standard-scale numeric features and one-hot encode categorical features with \texttt{handle\_unknown=`ignore'}. Preprocessed data is shared across all candidate pipelines for a fair comparison.
% ── Rule-Based Model Selection ────────────────────────────────────────────────
\subsection{Rule-Based Model Selection}

The \texttt{RuleBasedModelSelector} mechanism for exhaustive algorithm search is solved by providing rules based on metadata that relate to the size, complexity, class imbalance and cardinality of the data, its density of outliers, and its distributional skewness. Sample count $n$ is binned into five tiers: \emph{tiny} ($<$1\,K), \emph{small} (1\,K--10\,K), \emph{medium} (10\,K--100\,K), \emph{large} (100\,K--1\,M), and \emph{huge} ($\geq$1,M). A complexity score is derived from: high dimensionality ($p>50$, $+2$), high $p/n$ ratio ($>0.1$, $+2$), categorical dominance ($>50\%$, $+1$), and moderately many features ($p>20$, $+1$), mapping to \emph{simple} ($\leq 2$), \emph{moderate} (3--4), or \emph{complex} ($\geq 5$). The size-tiered base model pools for both types of task are shown in Table~\ref{tab:model_rules}.

\begin{table}[t]
\centering
\caption{Model pool by dataset-size tier. Models are selected based on
dataset size and complexity; see text for modifier rules.}
\label{tab:model_rules}
\setlength{\tabcolsep}{4pt}
\small
\begin{tabular}{@{}lp{4.0cm}p{4.0cm}@{}}
\toprule
\textbf{Tier} & \textbf{Classification} & \textbf{Regression} \\
\midrule
Tiny        & LR, DT, RF, NB, Ridge-LR, KNN         & OLS, Ridge, DT, Lasso, ElasticNet, KNN \\[2pt]
Small--Med. & RF, XGB, LGBM, GB, LR, KNN, NB        & RF, XGB, LGBM, GB, Ridge, KNN, SVR \\[2pt]
Large--Huge & LGBM, XGB, GB, LR(SAG), RF, Ridge-LR  & LGBM, XGB, GB, Ridge(SAG), RF, Lasso, SVR \\
\bottomrule
\multicolumn{3}{@{}l}{\scriptsize LR=Logistic Regression, DT=Decision Tree, RF=Random Forest,}\\
\multicolumn{3}{@{}l}{\scriptsize NB=Naive Bayes, KNN=$k$-NN, GB=Gradient Boosting,}\\
\multicolumn{3}{@{}l}{\scriptsize XGB=XGBoost, LGBM=LightGBM, SVR=Support Vector Regressor.}
\end{tabular}
\end{table}

Once the initial list is compiled the modifiers re-rank and add modifiers to the primary list: high dimensionality ($p>50$ or $p/n>0.1$), removes RF, SVM and adds regularised LR (L1/L2); the \texttt{is\_highly\_imbalanced} flag adds the \texttt{class\_weight=`balanced'}; high cardinality categoricals add tree-based models (XGB, LGBM, RF); outlier density~${>}5\%$ adds tree and demotes linear (after increasing precision of tree); skewed targets ($|\gamma_y|>1$), adds tree models (to make them independent of monotone transformation on target).  Each recommendation includes a priority ranking, a human-readable description and a hyperparameter (hp) grid, with the top $k=5$ passed on to pipeline generation.

% ── Training, Evaluation, Selection ──────────────────────────────────────────
\subsection{Training, Evaluation, and Selection}

The data is split 80/20, using stratified splitting for classification to maintain target distributions. A grid search (\texttt{GridSearchCV}) of a prescribed parameter space with 5-fold cross-validation (CV) on the training set is used to find the optimal set of hyperparameters. For example, the search space for Random Forest is \texttt{n\_estimators} $\in \{50, 100, 200\}$ and \texttt{max\_depth} $\in \{5, 10, \text{None}\}$, while for the gradient-boosting algorithms (XGBoost and LightGBM) it is \texttt{learning\_rate} $\in \{0.01, 0.1\}$ and \texttt{max\_depth} $\in \{3, 6\}$. The best-performing configuration is determined by mean CV performance (accuracy for classification \& $R^2$ for regression) with a refit on the entire training split (then tested on the test split). For classification, the metrics used for evaluation are accuracy, precision, recall, and weighted F1 score. For regression, metrics used for evaluation are $R^2$, Mean Squared Error (MSE), Root Mean Squared Error (RMSE), or Mean Absolute Error (MAE). In each of the following subsections, we examine the performance with different examples and make sure that the performance warning is triggered whenever needed (e.g., if the test score falls below 0.70, the train--test gap exceeds 0.10, or the cross-validation standard deviation $\sigma_{CV}$ exceeds 0.15). The most suitable pipeline is determined from the top-scoring test metric, and optimisation suggestions are given with issues identified.

% ══════════════════════════════════════════════════════════════════════════════
\section{Results}
\label{sec:results}

Evaluation of MetaFlow was conducted on four real-world datasets:
\textbf{D1 (Heart Disease)}~\cite{janosi1988heart} represents a small classification task ($n=1{,}025$, $p=13$, balanced targets with 51.3\% cases). Preprocessing uses median imputation, mode imputation, and scaling.
\textbf{D2 (Credit Fraud)}~\cite{dalpozzolo2015fraud} is a large, highly imbalanced classification task ($n=284{,}807$, $p=30$, 0.17\% minority class). Preprocessing scales the time and amount features.
\textbf{D3 (Brain Cancer)}~\cite{feltes2019cumida} is a high-dimensional multiclass task ($n=130$, $p=54{,}676$ expression features across 5 classes). Features are filtered to the top 1,000 by variance.
\textbf{D4 (House Sales)}~\cite{harlfoxem2016kc} is a regression task ($n=21{,}613$, $p=19$). Preprocessing includes log-transforms and one-hot encoding.

Table~\ref{tab:metadata_heatmap} summarises the key metadata properties;
highlighted cells indicate conditions that trigger modifier rules.

\begin{table}[t]
\centering
\caption{Metadata heatmap for D1--D4. Shading: \colorbox{hBlue}{\,blue\,}=minimal,
\colorbox{hGreen}{\,green\,}=low, \colorbox{hYellow}{\,yellow\,}=moderate,
\colorbox{hOrange}{\,orange\,}=high, \colorbox{hRed}{\,red\,}=critical.
Orange and red cells trigger modifier rules. Values are rounded for display; see
the repository for exact numbers.}
\label{tab:metadata_heatmap}
\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{@{}l cccc@{}}
\toprule
\textbf{Property} & \textbf{D1 Heart} & \textbf{D2 Fraud} & \textbf{D3 Brain} & \textbf{D4 House} \\
\midrule
Samples $n$      & \cellcolor{hYellow}1,025    & \cellcolor{hRed}284,807   & \cellcolor{hBlue}130      & \cellcolor{hOrange}21,613 \\
Features $p$     & \cellcolor{hGreen}13        & \cellcolor{hGreen}30      & \cellcolor{hRed}54,676    & \cellcolor{hGreen}19      \\
$p/n$ ratio      & \cellcolor{hGreen}0.013     & \cellcolor{hGreen}0.000   & \cellcolor{hRed}421.4     & \cellcolor{hGreen}0.001   \\
Size tier        & \cellcolor{hYellow}Small    & \cellcolor{hRed}Large     & \cellcolor{hBlue}Tiny     & \cellcolor{hOrange}Medium \\
Missing ratio    & \cellcolor{hGreen}0\%       & \cellcolor{hGreen}0\%     & \cellcolor{hGreen}0\%     & \cellcolor{hGreen}0\%     \\
Outlier density  & \cellcolor{hYellow}Mod.     & \cellcolor{hGreen}Low     & \cellcolor{hGreen}Low     & \cellcolor{hYellow}Mod.   \\
Class imbalance  & \cellcolor{hGreen}None      & \cellcolor{hRed}0.17\%    & \cellcolor{hGreen}None    & \cellcolor{hGray}N/A      \\
Target skewness  & \cellcolor{hGreen}Low       & \cellcolor{hGreen}Low     & \cellcolor{hGreen}Low     & \cellcolor{hOrange}High   \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Results Summary.}
MetaFlow selected sensible model families for each metadata regime: tree
ensembles for the large, imbalanced, and nonlinear cases; lightweight linear and
tree models for the tiny, high-dimensional case; and regression ensembles for
house-price prediction.  The detailed scores are reported in
Table~\ref{tab:all_results}.

\begin{table}[t]
\centering
\caption{Model performance across all four datasets.}
\label{tab:all_results}
\small
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.1}
\begin{tabular}{@{}llccc@{}}
\toprule
\textbf{Dataset} & \textbf{Model} & \textbf{CV Score} & \textbf{Test Score} & \textbf{Metric} \\
\midrule
\multirow{3}{*}{D1 Heart}
  & Random Forest        & 0.9707          & 0.9910          & Accuracy \\
  & XGBoost              & 0.9683          & 0.9920          & Accuracy \\
  & LightGBM             & \textbf{0.9720} & 0.9940          & Accuracy \\
\midrule
\multirow{3}{*}{D2 Fraud}
  & LightGBM             & \textbf{0.9983} & 0.9977          & Accuracy \\
  & XGBoost              & 0.9982          & 0.9977          & Accuracy \\
  & Random Forest        & 0.9982          & 0.9975          & Accuracy \\
\midrule
\multirow{3}{*}{D3 Brain}
  & Logistic Regression  & \textbf{0.9714} & 0.9231          & Accuracy \\
  & Gaussian Naive Bayes & 0.9137          & 0.9231          & Accuracy \\
  & Decision Tree        & 0.9042          & \textbf{0.9615} & Accuracy \\
\midrule
\multirow{3}{*}{D4 House}
  & Random Forest        & 0.8228          & 0.8442          & $R^2$ \\
  & XGBoost              & \textbf{0.8546} & \textbf{0.8453} & $R^2$ \\
  & LightGBM             & 0.8617          & 0.8401          & $R^2$ \\
\bottomrule
\multicolumn{5}{@{}l}{\scriptsize Bold = best score per column per dataset.}\\
\multicolumn{5}{@{}l}{\scriptsize D2: SVM and $k$-NN excluded ($O(N^2)$); \texttt{class\_weight=balanced} applied.}\\
\multicolumn{5}{@{}l}{\scriptsize D3: small-dataset rules selected lightweight models.}
\end{tabular}
\end{table}

% ══════════════════════════════════════════════════════════════════════════════
\section{Discussion}
\label{sec:discussion}

The four case studies answered both research questions: metadata alone was enough to guide top-level task and model choices, and the pruning strategy consistently reduced the candidate pool from nine models to three without sacrificing performance. The main limitation is the reliance on preset rule thresholds (like $\tau_\text{cat}=10$ and $\theta_\text{miss}=0.5$). While these are interpretable, they are dataset-dependent. Finally, because this is a very small dataset, the slightly lower scores should be seen as indicative rather than definitive.

% ══════════════════════════════════════════════════════════════════════════════
\section{Complexity, Limitations, and Deployment}
\label{sec:limitations}

To assess the practical viability of MetaFlow, this section presents a computational complexity analysis, discusses its limitations, and outlines deployment challenges.\subsection{Computational Complexity Analysis}MetaFlow operates in three distinct phases. First, \textbf{metadata extraction} requires $O(N \cdot P)$ execution time and $O(P)$ memory to compute statistical metrics over $N$ samples and $P$ features, completing in seconds. Second, \textbf{rule evaluation} executes in $O(1)$ time and memory. Third, by pruning the candidate pool from $M$ down to $k$ models ($k \ll M$), the \textbf{training and tuning} complexity is reduced from $O(M \cdot H \cdot f \cdot T_{\text{train}})$ to $O(k \cdot H \cdot f \cdot T_{\text{train}})$, yielding a $\sim$67\% training speedup. Finally, \textbf{inference complexity} is kept highly efficient by avoiding models with quadratic inference costs (e.g., $k$-NN, which requires $O(N \cdot P)$ per query) in favor of tree-based or linear structures that cost only $O(P)$.\subsection{Limitations and Deployment Challenges}MetaFlow faces several limitations and hurdles in real-world deployment. First, the selection rules rely on \textbf{heuristic thresholds} (e.g., a 5\% cutoff for class imbalance), which may require domain-specific calibration. Second, the system is designed strictly for \textbf{tabular data} and cannot process unstructured modalities like images or text. Third, \textbf{deployment in CI/CD pipelines} is complicated by schema drift and shifting target distributions. Because evolving dataset profiles will dynamically alter model recommendations, robust monitoring is essential to prevent performance regressions.\subsection{Benchmarking and Reproducibility}This study focuses on representative case studies rather than a comprehensive benchmark. A rigorous follow-up comparison should use identical data splits, nested cross-validation, and matched time budgets, reporting both predictive performance and evaluation costs. Complete reproduction commands and scripts are provided in REPRODUCIBILITY.md.

% ══════════════════════════════════════════════════════════════════════════════
\section{Conclusion}
\label{sec:conclusion}

This work presents an AutoML framework called \textbf{MetaFlow}, which leverages dataset metadata such as size, dimensionality, imbalance, skewness, and cardinality alongside heuristic rules to deterministically narrow the search space. Experiments across four real-world datasets, covering various task types (binary classification, highly imbalanced classification, high-dimensional multiclass classification, and regression), demonstrate that this ``grey-box'' approach reduces model evaluations by $\sim$67\% while maintaining highly competitive predictive performance.The core idea is that treating metadata as hard constraints, rather than just initialization hints, allows the system to eliminate unsuitable candidates \emph{before} training begins. This strikes an effective balance between a broad automated search and expert heuristics. Future work will expand this rule set using learned or LLM-assisted metadata interpretation.

\section*{Acknowledgement and Code}

The authors thank the Charotar University of Science and Technology (\href{https://www.charusat.ac.in/}{\textbf{CHARUSAT}}), Changa, India for support. The project codebase is available at \url{https://github.com/harsh-kakadiya1/AI-Agent-for-designing-ML-pipeline}.

\section*{}
% ══════════════════════════════════════════════════════════════════════════════
\begingroup
\let\clearpage\relax
\bibliographystyle{splncs04}
\setlength{\bibsep}{1.5pt plus 0.3pt}
\bibliography{references}
\endgroup

\end{document}